Gsmodutils: A python based framework for test-driven genome scale metabolic model development

### 2018-09-28 article/article.tex
Article ready for bioArxiv submission
Minor whitespace change
Submission copy

\begin{abstract}
\textbf{Motivation:}
Genome scale metabolic models (GSMMs) are increasingly important for systems biology and metabolic engineering research as they are capable of simulating complex steady-state behaviour.
Constraints based models of this form can include thousands of reactions and metabolites, with many crucial pathways that only become activated in specific simulation settings.
However, despite their widespread use, power and the availability of tools to aid with the construction and analysis of large scale models, little methodology is suggested for the continued management of curated large scale models.
For example, when genome annotations are updated or new understanding regarding behaviour of is discovered, models often need to be altered to reflect this.
This is quickly becoming an issue for industrial systems and synthetic biotechnology applications, which require good quality reusable models integral to the design, build and test cycle.
\\
\textbf{Results:}
As part of an ongoing effort to improve genome scale metabolic analysis, we have developed a test-driven development methodology for the continuous integration of validation data from different sources.
Contributing to the open source technology based around COBRApy, we have developed the \textit{gsmodutils} modelling framework placing an emphasis on test-driven design of models through defined test cases.
Crucially, different conditions are configurable allowing users to examine how different designs or curation impact a wide range of system behaviours, minimising error between model versions.\\
\textbf{Availability:} The software framework described within this paper is open source and freely available from \url{http://github.com/SBRCNottingham/gsmodutils} \\
\textbf{Contact:}{james.gilbert@nottingham.ac.uk}
\end{abstract}

@@

\section{Case study usage Clostridium autoethanogenum}
\textit{Clostridium autoethanogenum} is a bacterial species used for the production of commodity chemicals at industrial scale \cite{abrini1994clostridium,norman2018progress,LT}. A new GSMM of \textit{C. autoethanogenum}, `MetaCLAU', has been analysed to improve this bioprocess (Norman et al., submitted for peer review).
In this section we describe how \textit{gsmodutils} has been utilised to ensure that future versions of MetaCLAU will remain functionally relevant from the perspective of industrial biotechnology.

@@

\textit{C. autoethanogenum} is a strictly anaerobic, acetogenic bacterium which naturally produces ethanol and trace amounts of \textit{2,3-butanediol} (\textit{2,3 BD}) from carbon monoxide and water \cite{abrini1994clostridium,schuchmann2014autotrophy,norman2018progress}. Since carbon monoxide is readily available in the form of industrial waste gas, and \textit{2,3 BD} has a global market value of \$43 billion \cite{kopke201123butanediol}, the optimisation of yields of \textit{2,3 BD} from carbon monoxide is highly desirable in the context of industry \cite{norman2018progress}.   
MetaCLAU was built using Pathway Tools \cite{karp2002pathway} and ScrumPy \cite{poolman2006scrumpy}, and is based on a manually annotated genome sequence of \textit{C. autoethanogenum} \cite{humphreys2015whole}. The resulting model consisted of 758 reactions, 773 metabolites and 518 genes. For full details of the model, see (Norman et al. submitted for peer review).
The model has been integrated with the \textit{gsmodutils} modelling framework as a test-driven project. 
The following section details specific tests used to evaluate the model at each stage of its continued development.

@@

\textbf{Minimisation of flux based tests:}
One conventional approach in FBA is to set an optimisation criterion of minimising flux across enzymatic reactions with a fixed biomass constraint \cite{holzhutter2006thegeneralised}. 
The solution to this FBA problem represents minimal protein investment \cite{holzhutter2006thegeneralised}.
Since execution of flux minimisation in COBRApy requires a model in which reversible reactions are split into two irreversible reactions, a \textit{gsmodutils} design was created in which all reversible reactions are split in a programmatic manner.
In the case of MetaCLAU, the flux minimal solution includes both ethanol and acetate production, which represents good qualitative agreement with experimental data (Norman et al., unpublished).
Since any changes to this predicted phenotype must be investigated, the flux minimisation analysis has been formulated as a \textit{gsmodutils} test.

\textbf{Product scans:}
Of interest to this project were changes in the product spectrum of \textit{C. autoethanogenum} under conditions where the organism can and cannot produce molecular hydrogen (with carbon monoxide as sole carbon and energy source).
The hypothesis tested in (Norman et al., unpublished), was that in the case where hydrogen production infeasible, alternative electron sinks like lactate and \textit{2,3 BD} would be produced.
As in the previous case, the model-predicted behaviour showing both lactate and \textit{2,3 BD} was deemed an important result, which model curators should be notified of if lost during model development.
Thus the analysis was built into a \textit{gsmodutils} test.

@@

These test cases are all fully reproducible and available as an example within the \textit{gsmodutils} software repository and we encourage the reader to explore these examples to understand how to utilise \textit{gsmodutils} within their own project.

>>>>

\begin{abstract}
\textbf{Motivation:} Genome scale metabolic models (GSMMs) are increasingly important for systems biology and metabolic engineering research as they are capable of simulating complex steady-state behaviour.
Constraints based models of this form can include thousands of reactions and metabolites, with many crucial pathways that only become activated in specific simulation settings.
However, despite their widespread use, power and the availability of tools to aid with the construction and analysis of large scale models, little methodology is suggested for the continued management of curated large scale models.
For example, when genome annotations are updated or new understanding regarding behaviour of is discovered, models often need to be altered to reflect this.
This is quickly becoming an issue for industrial systems and synthetic biotechnology applications, which require good quality reusable models integral to the design, build and test cycle.
\\
\textbf{Results:}
As part of an ongoing effort to improve genome scale metabolic analysis, we have developed a test-driven development methodology for the continuous integration of validation data from different sources.
Contributing to the open source technology based around COBRApy, we have developed the \textit{gsmodutils} modelling framework placing an emphasis on test-driven design of models through defined test cases.
Crucially, different conditions are configurable allowing users to examine how different designs or curation impact a wide range of system behaviours, minimising error between model versions.\\
\textbf{Availability:} The software framework described within this paper is open source and freely available from \url{http://github.com/SBRCNottingham/gsmodutils} \\
\end{abstract}

@@

\section{Case study usage Clostridium autoethanogenum}
\textit{Clostridium autoethanogenum} is a bacterial species used for the production of commodity chemicals at industrial scale \cite{abrini1994clostridium,norman2018progress,LT}. A new GSMM of \textit{C. autoethanogenum}, `MetaCLAU', has been analysed to improve this bioprocess (\cite{norman2018progress}, submitted for peer review).
In this section we describe how \textit{gsmodutils} has been utilised to ensure that future versions of MetaCLAU will remain functionally relevant from the perspective of industrial biotechnology.

@@

\textit{C. autoethanogenum} is a strictly anaerobic, acetogenic bacterium which naturally produces ethanol and trace amounts of \textit{2,3-butanediol} (\textit{2,3 BD}) from carbon monoxide and water \cite{abrini1994clostridium,schuchmann2014autotrophy,norman2018progress}. Since carbon monoxide is readily available in the form of industrial waste gas, and \textit{2,3 BD} has a global market value of \$43 billion \cite{kopke201123butanediol}, the optimisation of yields of \textit{2,3 BD} from carbon monoxide is highly desirable in the context of industry \cite{norman2018progress}.   
MetaCLAU was built using Pathway Tools \cite{karp2002pathway} and ScrumPy \cite{poolman2006scrumpy}, and is based on a manually annotated genome sequence of \textit{C. autoethanogenum} \cite{humphreys2015whole}. The resulting model consisted of 758 reactions, 773 metabolites and 518 genes. For full details of the model, see (\cite{norman2018progress}. submitted for peer review).
The model has been integrated with the \textit{gsmodutils} modelling framework as a test-driven project. 
The following section details specific tests used to evaluate the model at each stage of its continued development.

@@

\textbf{Minimisation of flux based tests:}
One conventional approach in FBA is to set an optimisation criterion of minimising flux across enzymatic reactions with a fixed biomass constraint \cite{holzhutter2006thegeneralised}. 
The solution to this FBA problem represents minimal protein investment \cite{holzhutter2006thegeneralised}.
Since execution of flux minimisation in COBRApy requires a model in which reversible reactions are split into two irreversible reactions, a \textit{gsmodutils} design was created in which all reversible reactions are split in a programmatic manner.
In the case of MetaCLAU, the flux minimal solution includes both ethanol and acetate production, which represents good qualitative agreement with experimental data (\cite{norman2018progress}, under review).
Since any changes to this predicted phenotype must be investigated, the flux minimisation analysis has been formulated as a \textit{gsmodutils} test.

\textbf{Product scans:}
Of interest to this project were changes in the product spectrum of \textit{C. autoethanogenum} under conditions where the organism can and cannot produce molecular hydrogen (with carbon monoxide as sole carbon and energy source).
The hypothesis tested in (\cite{norman2018progress}, under review), was that in the case where hydrogen production infeasible, alternative electron sinks like lactate and \textit{2,3 BD} would be produced.
As in the previous case, the model-predicted behaviour showing both lactate and \textit{2,3 BD} was deemed an important result, which model curators should be notified of if lost during model development.
Thus the analysis was built into a \textit{gsmodutils} test.

@@
